---
title: "Preliminary analyses for considering the Right to Health in the Universal Periodic Review"
author: "Anshu Uppal"
date: "2025-06-03"
date-modified: last-modified # Dynamic modified date
format:
        html:
                code-fold: true
                toc: true
                toc-location: left  # Place TOC on the left side
                toc-depth: 5        # Include headers up to level 3 (###)
                toc-title: "On this page" # Custom title for the TOC
                smooth-scroll: true # Enable smooth scrolling
                embed-resources: true # Make the html file self contained
reference-location: margin
citation-location: margin
---

## Setup
```{r}
pacman::p_load(
  here,
  tidyverse,
  janitor,
  DT
)

# Load or install packages from GitHub:
pacman::p_load_gh(
  "DrMattG/SDGsR", # Uses API to get SDGs data
  "aphp/rgho" # Uses API to get data from Global Health Observatory
)
```

Read in the UPR data from the SDG data from the "Human Rights Data Explorer" (created and maintained by the Danish Institute for Human Rights):
<https://sdgdata.humanrights.dk/en/mechanisms/upr/cycle-4>
(Navigate to the link above and click "download all data" - the download preparation takes >30 minutes)

I then converted the CSV file to RDS format to save space:
```{r}
# Download from https://sdgdata.humanrights.dk/en/solr-explorer
sdg_data <- readRDS(here("data","sdg_data.rds")) |> 
  janitor::clean_names() |> 
  mutate_if(is.character, ~na_if(., "")) |> # convert blank spaces to true NA
  filter(mechanism == "Universal Periodic Review") |> 
  rename(state_under_review = state) |> 
  mutate(
    year = ymd(paste0(year, "01-01")),
    cycle = factor(cycle),
    sdg_linked = factor(case_when(sdg_goals == "No SDG link identified" ~ "No SDG link",
                                  .default = "Linked to an SDG"), 
                        levels = c("No SDG link", "Linked to an SDG")),
    response_upr = factor(response_upr),
    title = str_split_i(paragraph, "\\|", 2)
  ) |> 
  relocate(sdg_linked, .after = sdg_goals)

# Alternatively download straight from OHCHR website
# https://uhri.ohchr.org/en/our-data-api 
# OHCHR <- jsonlite::read_json(here("data", "export-full-en.json"))
# OHCHR <- tidyjson::read_json(here("data", "export-full-en.json"))
```

### Create definitions of right to health
I then created a set of keywords to assign recommendations as being related to the "right to health". This is a key step to refine in case we opt for an automated process. Some of the keywords are partial words because I use a partial word matching (e.g. "prophyla" will match with both "prophylaxis" as well as "prophylactics"). **Need to be careful to not accidentally include irrelevent terms.**
```{r}
# Define keywords to use to identify recommendations related to the right to health
health_keywords <- "health|well-being|wellbeing|sanitation|sanitary|safe water|clean water|right to water|disease|sick|infectio|nutrition|TB|tuberculosis|malaria|HIV|hospital|clinic|vaccin|immunisation|immunization|virus|viral|medic|doctor|nurse|nursing|maternal|contracep|abortion|reproductive|pregnan|prenatal|postnatal|neonatal|breasfe|obstetric|fertility|medic|diabetes|cancer|blood|hypertension|respiratory|prophyla|violence against women|violence against children|domestic violence|gender-based violence|sexual violence|marital violence|gender violence|virginity|hunger|malnutrition|obesity|sexual abuse|mortality|leprosy|famine|drought|food|hygiene|tobacco|smoking|child marriage|forced marriage|infanticide|mutilation|honor crime|honor killing|honour crime|honour killing|breast ironing|harmful practices|harmful traditional practices|disaster|disabled|disabili"

# Define combinations of keywords for the right to health
keywords_comb1a <- "child|girl|women|sexual|domestic|gender|marital"
keywords_comb1b <- "abuse|maltreatment|violence|sexual"

keywords_comb2a <- "forced"
keywords_comb2b <- "marriage"

keywords_comb3a <- "sex|gender|civil identity|transgender"
keywords_comb3b <- "surgery|sterili"
```

### Apply the definitions
```{r}
sdg_data <- sdg_data |> 
  mutate(
    health_related = factor(case_when(
      str_detect(tolower(text), health_keywords)| 
        str_detect(tolower(sdg_goals), "health|sanitation")
      # str_detect(targets, "5.6|16.1|16.2")
      ~ "Health-related",
      # Combo 1
      str_detect(tolower(text), keywords_comb1a)&
        str_detect(tolower(text), keywords_comb1b)~ "Health-related",
      # Combo 2
      str_detect(tolower(text), keywords_comb2a)&
        str_detect(tolower(text), keywords_comb2b)~ "Health-related",
      # Combo 3
      str_detect(tolower(text), keywords_comb3a)&
        str_detect(tolower(text), keywords_comb3b)~ "Health-related",
      
      .default = "Not health-related"
    ), levels = c("Not health-related", "Health-related"))
  ) |> 
  relocate(health_related, .after = text)
```

### Analysis
#### Percentage of recommendations linked to the right to health:
```{r}
sdg_data |> 
  group_by(cycle) |> 
  count(health_related) |> 
  mutate(perc = n/sum(n)*100) |> 
  ggplot(aes(x = cycle, y = n, fill = health_related, label = paste0(n, " (", round(perc,1), "%)")))+
  geom_col()+
  labs(y = "Number of recommendations", x = "UPR Cycle", 
       title = "UPR recommendations related to the right to health, per cycle",
       fill = "")+
  geom_text(position = position_stack(), vjust = 1)+
  theme_bw()
```

#### Percentage of recommendations linked to SDGs
```{r}
sdg_data |> 
  group_by(cycle, health_related) |> 
  count(sdg_linked) |> 
  mutate(perc = n/sum(n)*100) |> 
  ggplot(aes(x = cycle, y = n, fill = sdg_linked, label = paste0(round(perc,1), "%")))+
  facet_grid(~ health_related, 
             scales = "free_x", 
             space = "free_x", 
             switch = "x"
  )+
  geom_col()+
  labs(x = "Number of recommendations", x = "Cycle", fill = "",
       title = "UPR recommendations that have been linked to an SDG target")+
  geom_text(position = position_stack(), vjust = 1)+
  theme_bw()
```

#### State responses to UPR recommendations - by linkage to right to health
```{r}
sdg_data |> 
  filter(!is.na(response_upr)) |> 
  group_by(cycle, health_related) |> 
  count(response_upr) |> 
  mutate(perc = n/sum(n)*100) |> 
  ggplot(aes(x = cycle, y = n, fill = response_upr, label = paste0(round(perc,1), "%")))+
  facet_grid(~ health_related, 
             scales = "free_x", 
             space = "free_x", 
             switch = "x"
  )+
  geom_col()+
  labs(x = "Number of recommendations", x = "Cycle", fill = "",
       title = "State responses to UPR recommendations")+
  geom_text(position = position_stack(), vjust = -0.5)+
  theme_bw()
```

#### State responses to UPR recommendations - by SDG linkage
```{r}
sdg_data |> 
  filter(!is.na(response_upr)) |> 
  group_by(cycle, sdg_linked) |> 
  count(response_upr) |> 
  mutate(perc = n/sum(n)*100) |> 
  ggplot(aes(x = cycle, y = n, fill = response_upr, label = paste0(round(perc,1), "%")))+
  facet_grid(~ sdg_linked, 
             scales = "free_x", 
             space = "free_x", 
             switch = "x"
  )+
  geom_col()+
  labs(x = "Number of recommendations", x = "Cycle", fill = "",
       title = "State responses to UPR recommendations")+
  geom_text(position = position_stack(), vjust = -0.5)+
  theme_bw()
```

#### State responses to UPR recommendations - Overall
```{r}
sdg_data |> 
  filter(!is.na(response_upr)) |> 
  group_by(cycle) |> 
  count(response_upr) |> 
  mutate(perc = n/sum(n)*100) |> 
  ggplot(aes(x = cycle, y = n, fill = response_upr, label = paste0(round(perc,1), "%")))+
  geom_col(position = "dodge")+
  labs(y = "Number of recommendations", x = "UPR Cycle", 
       title = "State responses to UPR recommendations",
       fill = "")+
  geom_text(position = position_dodge(width = 0.9), vjust = -0.5)+
  theme_bw()
```

#### Further explore the assigned recommendations

I've split the recommendations into two groups here, one where they've been linked to the right to health and one where the link has not been made. These can be explored and inspected, for example by adding in keywords into the filters, so that we can make sure that we aren't missing or inappropriately including recommendations within our definitions for the "right to health".
```{css}
/* set the DT table fontsizes */
th { font-size: 11px; } /* header font */
td { font-size: 11px; } /* cell font */
```

##### Linked to right to health:
```{r}
#| warning: false

yes_health <- sdg_data |> 
  mutate(state_under_review = factor(state_under_review)) |> 
  filter(health_related == "Health-related") |> 
  select(text, sdg_goals,
         # groups, 
         state_under_review, response_upr, cycle, document_code, paragraph)

yes_health |> 
  DT::datatable(
    filter = "top",
    options = list(
      pageLength = 100,
      deferRender = TRUE,
      scrollY = 400,
      scrollX = TRUE,
      scroller = TRUE,
      autoWidth = TRUE,      
      columnDefs = list(list(width = '500px', targets = c(0)),
                        list(width = '200px', targets = c(1)))
    ),
    rownames = FALSE, # set to FALSE for cleaner look
    class = 'cell-border stripe hover compact'
  )

```

###### Random sample
```{r}
yes_health |> slice_sample(n = 500) |> 
  DT::datatable(
    filter = "top",
    options = list(
      pageLength = 100,
      deferRender = TRUE,
      scrollY = 400,
      scrollX = TRUE,
      scroller = TRUE,
      autoWidth = TRUE,      
      columnDefs = list(list(width = '500px', targets = c(0)),
                        list(width = '200px', targets = c(1)))
    ),
    rownames = FALSE, # set to FALSE for cleaner look
    class = 'cell-border stripe hover compact'
  )
```

##### Not linked to right to health:
```{r}
#| warning: false

not_health <- sdg_data |> 
  mutate(state_under_review = factor(state_under_review)) |> 
  filter(health_related == "Not health-related") |> 
  select(text, sdg_goals, 
         # groups, 
         state_under_review, cycle, document_code, paragraph)

not_health |> 
  DT::datatable(
    filter = "top",
    options = list(
      pageLength = 100,
      deferRender = TRUE,
      scrollY = 400,
      scrollX = TRUE,
      scroller = TRUE,
      autoWidth = TRUE,      
      columnDefs = list(list(width = '500px', targets = c(0)),
                        list(width = '200px', targets = c(1)))
    ),
    rownames = FALSE, # set to FALSE for cleaner look
    class = 'cell-border stripe hover compact'
  )
```

###### Random sample
```{r}
not_health |> slice_sample(n = 500) |> 
  DT::datatable(
    filter = "top",
    options = list(
      pageLength = 100,
      deferRender = TRUE,
      scrollY = 400,
      scrollX = TRUE,
      scroller = TRUE,
      autoWidth = TRUE,      
      columnDefs = list(list(width = '500px', targets = c(0)),
                        list(width = '200px', targets = c(1)))
    ),
    rownames = FALSE, # set to FALSE for cleaner look
    class = 'cell-border stripe hover compact'
  )
```

## Focus on Cycle 3

### Load in dataset from UPR Info
I downloaded the UPR Info dataset on 2025-06-03 as several individual files (the download size was limited to 10,000 recommendations per download) and then combined them into one main file.

**(Still to be worked on)**
```{r}
# upr_info <- readRDS(here("data", "UPRInfo_all.rds")) |> 
#   janitor::clean_names() |>
#   mutate(
#     state_under_review = str_split_i(related_document, " - ", 1),
#     cycle = factor(str_split_i(cycle, " \\(", 1))
#   )

# upr_info_cycle_3 <- upr_info |> 
#   filter(cycle == "Cycle 3") |> droplevels() |> 
#   select(title,cycle, state_under_review, recommendation, action_category, issues) |> 
#   left_join(
#     sdg_data |> filter(cycle == "Cycle 3") |> droplevels(), 
#     join_by(state_under_review, title)
#     )
```

## Link to Indicators from GHO

Make API calls to the GHO database to download the data related to the UHC indicators and to the indicator for maternal mortality ratio
```{r}
#| message: false

## Metadata ####
gho_codes <- get_gho_values(dimension = "GHO") |> 
  rename(IndicatorCode=Code, indicator_fullname = Title)
country_codes <- get_gho_values(dimension = "COUNTRY") |> 
  rename(COUNTRY = Code, country_name = Title)
region_codes <- get_gho_values(dimension = "REGION") |> 
  rename(REGION = Code, region_name = Title)

# Get the UHC data
UHC_AVAILABILITY_SCORE <- get_gho_data(code = "UHC_AVAILABILITY_SCORE")
UHC_INDEX_REPORTED <- get_gho_data(code = "UHC_INDEX_REPORTED")
UHC_SCI_CAPACITY <- get_gho_data(code = "UHC_SCI_CAPACITY")
UHC_SCI_INFECT <- get_gho_data(code = "UHC_SCI_INFECT")
UHC_SCI_NCD <- get_gho_data(code = "UHC_SCI_NCD")
UHC_SCI_RMNCH <- get_gho_data(code = "UHC_SCI_RMNCH")

# Combine the datasets and tag indicator and country full names
UHC_all <- bind_rows(UHC_AVAILABILITY_SCORE, UHC_INDEX_REPORTED, 
                     UHC_SCI_CAPACITY, UHC_SCI_INFECT,
                     UHC_SCI_NCD, UHC_SCI_RMNCH) |> 
  left_join(country_codes) |> 
  left_join(gho_codes) |> 
  left_join(region_codes) |> 
  mutate(
    NumericValue = as.numeric(NumericValue),
    year = ymd(paste0(YEAR, "-01-01")),
    region_name = case_when(GLOBAL == "GLOBAL" ~ "Global", 
                            .default = region_name)
    )

# Maternal mortality ratio
MMR <- get_gho_data(code = "MDG_0000000026") |> 
  left_join(country_codes) |> 
  left_join(gho_codes) |> 
  left_join(region_codes) |> 
  mutate(
    NumericValue = as.numeric(NumericValue),
    year = ymd(paste0(YEAR, "-01-01")),
    region_name = case_when(GLOBAL == "GLOBAL" ~ "Global", 
                            .default = region_name)
  )
```

### Plots
#### UHC Indicators
```{r}
# Plot
UHC_all |> 
  filter(!IndicatorCode %in% c("UHC_AVAILABILITY_SCORE",
                               "UHC_INDEX_REPORTED")) |> 
  filter(country_name %in% c("India", "Namibia", "Rwanda", "Sierra Leone", 
                             "Sudan", "Argentina")) |> 
  mutate(indicator_fullname = str_to_sentence(
    str_remove(indicator_fullname, "UHC Service Coverage sub-index on ")
    )) |> 
  ggplot(aes(x=year, y = NumericValue, color = indicator_fullname, shape = indicator_fullname))+
  geom_point()+
  geom_line()+
  labs(y = "Index value",
       x = "Year",
       title = "UHC Service Coverage Sub-Indexes", color = "", shape = "")+
  facet_wrap(.~country_name)+
  theme_bw()+
  theme(legend.position = "bottom")+
  guides(color=guide_legend(nrow=2,byrow=TRUE))
```

#### Maternal mortality ratio
```{r}
# Plot
MMR |> 
  filter(country_name %in% c(
    "India", "Namibia", "Rwanda", "Sierra Leone",
                             "Sudan",
                             "Argentina")) |> 
  filter(year > ymd("1994-01-01")) |> 
  ggplot(aes(x=year, y = NumericValue))+
  geom_line(color = "tomato", size = 1)+
  labs(y = "Index value",
       x = "Year",
       title = "Maternal mortality ratio (per 100,000 live births)")+
  facet_wrap(.~country_name, scales = "free")+
  theme_bw()+
  theme(legend.position = "bottom")+
  guides(color=guide_legend(nrow=2,byrow=TRUE))

# Plot
MMR |> 
  filter(!is.na(region_name)) |> 
  filter(!region_name %in% c("Global")) |> 
  # filter(year > ymd("1994-01-01")) |> 
  ggplot(aes(x=year, y = NumericValue, color = region_name))+
  geom_line(size = 1)+
  labs(y = "Value",
       x = "Year",
       title = "Maternal mortality ratio (per 100,000 live births)",
       color = "")+
  # facet_wrap(.~region_name)+
  theme_bw()+
  theme(legend.position = "bottom")+
  guides(color=guide_legend(nrow=2,byrow=TRUE))

```

